Supplementary Data: 14C Modeling for Kiwulan Chronology

### 2020-08-22 analysis/paper/KWL-chronology-modeling.Rmd
update file, code, and text for the C14 modeling file

---
title: "Supplementary Data: 14C Modeling for Kiwualn Chronology"
author: "Liying Wang and Ben Marwick"
date: "8/18/2020"
output: html_document
bibliography: references.bib
---

@@

This document provides data and code for modeling the chronology of the upper component of Kiwulan. The full discussion of chronology is in the paper: The paper Standardization of ceramic shape: A case study from the Iron Age pottery from northeastern Taiwan. In this document, we calibrated all C14 dates [@Chen2007] first and then further modeled different phases related to foreign influences, pre- and post- European periods. Our purpose is to answer whether the C14 dates support a time range that corresponds to historical events of the presence of Europeans in Taiwan. The methods and results are included below.

@@

# make some kind of ID
kw_dates_with_error$id <- with(kw_dates_with_error,
                               paste0(Square,"-",Layer))
```

@@

We calibrated 23 dates from different contexts, including normal deposit, burials, and middens, at Kiwulan using oxcAAR package, which executes OxCal v4.4.1 with radiocarbon calibration curve IntCal20 [@Reimer2020]. <!-- if this is not the full set of ages in Chen, then please explain how you choose them -->

@@

```{r data-for-modeling}
# upper layer with estimation of phases
kw_dates_with_error_three_phases <-
  kw_dates_with_error %>%
  filter(rce < 600) %>%
  mutate(phase = rce) %>%
  mutate(phase = ifelse(rce > 324, "Pre European", phase),
         phase = ifelse(rce < 324, "Post European", phase)) %>%
  select(`Lab code`, "rce", "se", "phase")
```

@@

We then constructed a Bayesian radiocarbon calibration model to obtain the range of pre-European and post-European we examined. We assume the phases are in an order that represents before and after the European arrival in 1624 [@Andrade2007]. The end of post-European is identified by the start of the Chinese migration to Yilan in 1780s [@Chen2007].  <!-- please explain how you choose the subset of ages to go into this model -->

The results suggest that the re-European period started between 1331–1635 cal AD at a 95.4% probability range and ended between 1458–1670 cal AD at a 99.7% probability range. The end date of pre-European also represents the start date of post-European, while the end date of post-European is 1517-1753 cal AD at a 95.4% probability range. The overall model is consistent with the historical data.

```{r modeling-upper-layer}
# contiguous phases, not overlapping
three_phases_con <- 'Plot()
{
    Sequence("Upper Layer")
    {
      Boundary("Start of Pre-European");
      Phase()
      {
        R_Date("NTU-3791", 340, 30);
        R_Date("NTU-4292", 510, 75);
        R_Date("NTU-4310", 360, 100);
        R_Date("NTU-4320", 340, 100);
        Interval( "Duration of Pre-European");
      };
        Boundary("Pre-European - Post-European");
      Phase()
      {
        R_Date("NTU-3993", 250, 40);
        R_Date("NTU-4016", 270, 45);
        R_Date("NTU-4311", 310, 100);
        R_Date("NTU-4419", 280, 70);
        Interval("Duration of Post-European");
      };
      Boundary("End of Post-European");
    };
};
'
```

>>>>

---
title: "Supplementary Data: 14C Modeling for Kiwulan Chronology"
author: "Li-Ying Wang and Ben Marwick"
date: "8/18/2020"
output: html_document
bibliography: references.bib
---

@@

This document provides data and code for calibrating C14 dates and modeling the chronology of the upper component of Kiwulan. The full discussion of chronology is in the paper: Standardization of ceramic shape: A case study from the Iron Age pottery from northeastern Taiwan. There are 17 C14 dates from the upper component according to the excavation report [@Chen2007]. We calibrated 11 C14 dates after removing 6 dates that are not informative due to a long time range without standard deviation (< 200 yr). We further modeled two phases related to the European influence, pre- and post- European periods. Our purpose is to answer whether the C14 dates support a time range that corresponds to the historical events of the presence of the Europeans in Taiwan.

@@

# make some kind of ID
kw_dates_with_error$id <- with(kw_dates_with_error,
                               paste0(Square,"-",Layer))

# upper layer with estimation of phases
kw_dates_with_error_three_phases <-
  kw_dates_with_error %>%
  filter(Component == "upper") %>%
  mutate(phase = rce) %>%
  mutate(phase = ifelse(rce > 324, "Pre European", phase),
         phase = ifelse(rce < 324, "Post European", phase)) %>%
  select(`Lab code`, "rce", "se", "phase", Context)
```

@@

We calibrated 11 dates from different contexts, including artifact-bearing deposit, burials, and middens, at Kiwulan using oxcAAR package, which executes OxCal v4.4.1 with radiocarbon calibration curve IntCal20 [@Reimer2020].

@@

@@

We then constructed a Bayesian radiocarbon calibration model for the 11 dates to obtain the range of the pre-European and the post-European periods we examined. We assume that the two phases are in a time sequence with the arrival of the Europeans in 1624 as a break [@Andrade2007]. The end of the post-European period is identified by the start of the Chinese migration to Yilan in 1780s [@Chen2007].

The results suggest that the pre-European period started between 998–1386 cal AD at a 95.4% probability range and ended between 1496–1656 cal AD at a 95.4% probability range. The end date of pre-European is the same as the start date of post-European (1496–1656 cal AD). The end date of post-European is 1524-1770 cal AD at a 93.4% probability range. In general, the overall model is consistent with the historical data that we expect.

```{r modeling-upper-layer}
# contiguous phases, not overlapping
three_phases_con <- 'Plot()
{
    Sequence("Upper Layer")
    {
      Boundary("Start of Pre-European");
      Phase()
      {
        R_Date("NTU-3791", 340, 30);
        R_Date("NTU-4292", 510, 75);
        R_Date("NTU-4310", 360, 100);
        R_Date("NTU-4320", 340, 100);
        R_Date("NTU-4304", 600, 75);
        R_Date("NTU-4423", 610, 90);
        R_Date("NTU-4319", 920, 105);
        Interval( "Duration of Pre-European");
      };
        Boundary("Pre-European - Post-European");
      Phase()
      {
        R_Date("NTU-3993", 250, 40);
        R_Date("NTU-4016", 270, 45);
        R_Date("NTU-4311", 310, 100);
        R_Date("NTU-4419", 280, 70);
        Interval("Duration of Post-European");
      };
      Boundary("End of Post-European");
    };
};
'
```
